---
title: "Assignment 1"
author: "Yuhaniz Aly"
date: "4/23/2021"
output: html_document
---

```{r setup, include=FALSE}
source("analysis.R")
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```

## Describing Metrics

There are many different measurement of disease burden, which includes deaths,
YLLs, YLDs, DALYS.

**Years of Life Lost (YLLs)** measures the years of potential life lost due to
premature deaths.  To calculate YLLs for a particular outcome in a population,
we use the life expectancy for the given age minus the age, multiplied by the
number of deaths.

**Years Lived with Disability (YLDs)** measures the years of potential life lost
due to living with a disability. To calculate YLDs for a particular outcome in a
population, we use the disability weight, which is the magnitude of the
impairment, multiplied by the number of people living with that impairment in
that population (prevalence). 

**Disability Adjusted Life Years (DALYs)** is a measure of overall disease
burden, expressed the years of potential life lost due to ill-health,
disability, or premature death. DALYs combines mortality and morbidity into one
metric. To calculate DALYs for a particular outcome in a population, we take the
sum of the years life lost due to premature deaths (YLLs) and the years life
lost due to living with a disability (YLDs).

**Disability weights** quantifies the severity of outcomes as percentage
reduction from perfect health, which are multiplied by prevalent cases.
Disability weights is measured from a scale of 0 to 1, with 0 implying the
state in full health and 1 implying a state of death.

**Age-weighting** provides different values to life years lived at different
ages. Age-weighting assumes that a year of life increases in value until age 22,
and then decreases. I think age-weighting should be included in the calculations
of DALYs because I think young adulthood are when people are in their prime
years of health and less susceptible to contracting a disease that would put
them close to a near-death situation.

**Discounting** is an adjustment that reduces the burden of early mortality.
Discounting counts the years of healthy life saved in the present more valuable
than years of life saved in the future. I think discounting should not be
included in the calculations of DALYs because of the uncertainty correlating
with time. Some individuals might want to estimate their personal risk of death
each year at their own rate and preference, as it may vary from year to year,
rather than go with the default discount rate.

## Burden by Cause
This section compares the different measurements of health burden (deaths, YLLs,
YLDs, DALYs) across causes.

```{r plot2, echo = FALSE, warning=FALSE, fig.width = 12, fig.height = 6}
cause_patchwork
```
The barcharts shows the top 10 causes of the various metrics of health. Based on
the barcharts, _ischemic heart disease_ is the number one top cause of health 
burden, as it has the most number of deaths, YLLs, and DALYs. The barchart
illustrates that ischemic heart disease has significantly more cases of deaths,
YLLs, and DALYs than any other causes. In contrast, _low back and neck pain_ has
the most number of YLDs. In fact, ischemic heart disease did not appear to be
one of the top 10 causes for YLDs. This suggests that many people diagnosed with
ischemic heart disease tend to die from this disease and that many people lose
many years of their lifetime, which explains the large number of deaths, YLLs,
and DALYs. This means there will be fewer people living with ischemic heart
disease because they do not survive as long. This means there will be less cases
of people living with ischemic heart disease, which is why ischemic heart
disease is not a top 10 cause for YLDs. Meanwhile, low back and neck pain cases
are less severe on people's health and more common in the population, resulting
YLDs to be more prevalent.

## Burden by Age
This section compares the different measurements of health burden (deaths, YLLs,
YLDs, DALYs) across ages between 0 to 80 years old. Note: Children under 5 were
indicated as age 0 in the graphs.

```{r, echo = FALSE, warning=FALSE, fig.width = 12, fig.height = 6}
age_patchwork
```
The light blue barchart shows that the _number of deaths increases across
increasing age_, with people over 80 having the most deaths. On the other hand,
children under 5 have the second most deaths. This is not surprising because we
can expect an average person’s life expectancy to be around the 80s, though it
may slightly vary depending on geographic location. Also, children dying before
the age of 5 is unfortunately common across the world. This accounts for the
_large number of YLLs and DALYs in the under 5 age group_ because many children
lost so many years of their potential lifetime from an early death. The graphs
(light pink and light green barcharts) also show that in the later life, YLLs
and YLDs leveled off in older age groups. This means they do not have a high
life expectancy, resulting in a lower YLLs and DALYs value. On the other hand,
the light yellow barchart shows that
_the middle age groups, between  30 to 40 years old, tend to have the most YLDs_
.This is also not surprising because people in this age group tend to be
affected by health condition at this in their lifetime. The older groups
(65 years and older) have less YLDs because there not many people who reach this
age in life, which we can see in the light blue barchart.

## Burden by Sex
This section compares the different measurements of health burden (deaths, YLLs,
YLDs, DALYs) across sexes, males and females.

```{r, echo = FALSE, warning=FALSE, fig.width = 12, fig.height = 6}
sex_patchwork
```
Across three of the barcharts (light blue, light pink, and light green), they
illustrate that
_males have more number of deaths, YLLs, and DALYs than females_. Whereas in the
light yellow barchart, _females have more YLDs than males_. This suggest that
males die more than females and that they lose more years their lifetime from
the health cause/disease. This also means that more females live the disease or
disabilty than males and that females do not lose many years of their lifetime
from the cause than males. Additionally, this may suggest that there are more
males in the population affected by diseases than females because males die the
most.